---
title: "Untitled"
author: "Carsten Behring"
date: "Tuesday, October 21, 2014"
output: html_document
---

# Sope of Analysis
The scope of this analsys is to build a model in order to predict if a ertain weigh-lifting exercises (10 repetitions of the Unilateral Dumbbell Biceps Curl) is correctly executed or not.
The predictors are measurements of several sensors placed at the human body during the exercise.

For details, please see th original paper here: 
http://groupware.les.inf.puc-rio.br/public/papers/2013.Velloso.QAR-WLE.pdf


# Summary
By using sensor data which measures the 3D movements of the body at 4 specific body locations (arm, forearm, dumbbell, belt) it is possible to predict if a exercise was done correctly or not (and which mistake has been made)

Using the randomForest algorithm we get an OOB estimate for the error rate on the training data of 0.8 %. 



```{r echo=F,results='hide',warning=FALSE,message=F}
library(randomForest)
library(dplyr)
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}


```
## Data aquisition
The raw data is accessible here http://groupware.les.inf.puc-rio.br/har.

The data set contains the following information:

- "classe" : the result, if the exercise was executed correctly or not. It is encoded as:
     + A : correctly according to the specification 
     + B : throwing the elbows to the front
     + C : lifting the dumbbell only halfway
     + D : lowering the dumbbell only halfway
     + E : throwing the hips to the front
- several values about the time of the measurement
- raw sensor data
- summary statistics of sensor data per time window

Read more: http://groupware.les.inf.puc-rio.br/har#ixzz3H3Otr2Fy


```{r cache=TRUE,echo=FALSE,results='hide',warning=FALSE}

testing <- read.csv("./pml-testing.csv",stringsAsFactors=F) %>% select(-user_name,-new_window)
testing <- as.data.frame(lapply(testing,as.numeric))


training <- read.csv("./pml-training.csv",stringsAsFactors=F,na.strings=c("NA",""))
training <- training %>% select(-user_name,-new_window)
training <- training %>% mutate(classe=factor(training$classe))
classe <- training$classe;
training <- as.data.frame(lapply(training,as.numeric))
training$classe <- classe                                 



```

## Feature selection

I decided to use only the predictors with the raw sensor measurements and exclude therefore all others parameters.
Most of the excluded varaibles are summary statistcis of the raw data.

As they are summaries per time window, they should not add anything substantial new to the model.

So the final feature set used for training is this:

```{r}
predictors_to_keep <- c("classe",grep("_x$|_y$|_z$|^raw.*|num_window",names(training),value = T))
predictors_to_keep
```



```{r echo=FALSE}
training <- training[,predictors_to_keep]
testing <- testing[,predictors_to_keep[-1]]
```

We train then a model which is using "classe" as response. This variables encodes the outcome of the exercise with:

- correct (A)
- error made (B,C,D,E)

```{r cache=TRUE}
fit <- randomForest(classe ~ .,data=training)
fit
```

We see in the confusion matrixof the training process, that the estimated accuracy of the model for out-of-sample predictions values is about 99.2 % overall.
An explicit cross-validation is not needed by using the random forest algorithm, as it is part of teh algorithm already.


Prediction:

The predcition of teh values from teh test data set is as following:
```{r}
prediction <- predict(fit,newdata=testing)
prediction
pml_write_files(prediction)
```





